---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# tidytof: An ecosystem for tidy and highly-reproducible CyTOF data analysis

<!-- badges: start -->

<!-- badges: end -->

The goal of `tidytof` is to provide an integrated suite of R functions for CyTOF data analysis using an intuitive and internally consistent set of design principles. As an extension of the `tidyverse` ecosystem of data manipulation tools in R, `tidytof` is both performant and easy-to-use for scientists with a wide range of coding experience (including beginners).



## Installation

You can install the development version of tidytof from GitHub with the following command:

```{r, eval=FALSE}
if(!require(devtools)) install.packages("devtools") 
devtools::install_github("keyes-timothy/tidytof")
```


## Usage

### Reading data (including example datasets)

### Pre-processing 

### Clustering

### Dimensionality reduction

### Differential Expression Analysis

### Patient-level feature extraction

### Outcomes modeling

### Post-processing

### Writing data



<https://github.com/r-lib/actions/tree/master/examples>.


