---
title: "Dimensionality reduction"
output: 
  officedown::rdocx_document:
    mapstyles:
      Normal: ['First Paragraph']
    reference_docx: reference.docx

---

```{r setup_5, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE, 
  warning = FALSE,
  fig.cap = TRUE, 
  out.width = "100%"
  )
library(officedown)
library(officer)

fp <- fp_par(
  text.align = "center", 
  padding.bottom = 20, padding.top = 120, 
  border.bottom = fp_border())

ft <- fp_text(shading.color='#EFEFEF', bold = TRUE)

library(tidytof)
library(dplyr)
library(ggplot2)
```

A useful tool for visualizing the phenotypic relationships between single cells and clusters of cells is dimensionality reduction, a form of unsupervised machine learning used to represent high-dimensional datasets in a smaller number of dimensions.

`{tidytof}` includes several dimensionality reduction algorithms commonly used by biologists: Principal component analysis (PCA), t-distributed stochastic neighbor embedding (tSNE), and uniform manifold approximation and projection (UMAP). To apply these to a dataset, use `tof_reduce_dimensions()`. 

### Dimensionality reduction with `tof_reduce_dimensions()`. 

Here is an example call to `tof_reduce_dimensions()` in which we use tSNE to visualize data in `{tidytof}`'s built-in `phenograph_data` dataset.

```{r}
# perform the dimensionality reduction
phenograph_tsne <- 
  phenograph_data %>% 
  tof_preprocess() %>% 
  tof_reduce_dimensions(method = "tsne")

# select only the tsne embedding columns
phenograph_tsne %>% 
  select(contains("tsne")) %>% 
  head()
```

By default, `tof_reduce_dimensions` will add reduced-dimension feature embeddings to the input `tof_tbl` and return the augmented `tof_tbl` (that is, a `tof_tbl` with new columns for each embedding dimension) as its result. To return only the features embeddings themselves, set `add_cols` to `FALSE` (as in `tof_cluster`).

```{r}
phenograph_data %>% 
  tof_preprocess() %>% 
  tof_reduce_dimensions(method = "tsne", add_cols = FALSE) %>% 
  head()
```

Changing the `method` argument results in different low-dimensional embeddings: 

```{r}
phenograph_data %>% 
  tof_reduce_dimensions(method = "umap", add_cols = FALSE) %>% 
  head()

phenograph_data %>% 
  tof_reduce_dimensions(method = "pca", add_cols = FALSE) %>% 
  head()
```


### Method specifications for `tof_reduce_*()` functions

`tof_reduce_dimensions()` provides a high-level API for three lower-level functions: `tof_reduce_pca()`, `tof_reduce_umap()`, and `tof_reduce_tsne()`. The help files for each of these functions provide details about the algorithm-specific method specifications associated with each of these dimensionality reduction approaches. For example, `tof_reduce_pca` takes the `num_comp` argument to determine how many principal components should be returned: 

```{r}
# 2 principal components
phenograph_data %>% 
  tof_reduce_pca(num_comp = 2) %>% 
  head()
```

```{r}
# 3 principal components
phenograph_data %>% 
  tof_reduce_pca(num_comp = 3) %>% 
  head()
```

see `?tof_reduce_pca`, `?tof_reduce_umap`, and `?tof_reduce_tsne` for additional details.

### Visualization using `tof_plot_cells_dr()`

Regardless of the method used, reduced-dimension feature embeddings can be visualized using `{ggplot2}` (or any graphics package). `{tidytof}` also provides some helper functions for easily generating dimensionality reduction plots from a `tof_tbl` or tibble with columns representing embedding dimensions:

```{r}
# plot the tsne embeddings using color to distinguish between clusters
phenograph_tsne %>% 
  tof_plot_cells_dr(
    dr_cols = contains(".tsne"), 
    color_col = phenograph_cluster
  )

# plot the tsne embeddings using color to represent CD11b expression
phenograph_tsne %>% 
  tof_plot_cells_dr(
    dr_cols = contains(".tsne"), 
    color_col = cd11b
  ) + 
  ggplot2::scale_fill_viridis_c()
```

Such visualizations can be helpful in qualitatively describing the phenotypic differences between the clusters in a dataset. For example, in the example above, we can see that one of the clusters has high CD11b expression, whereas the others have lower CD11b expression.
