---
title: "Clustering and metaclustering"
author: "Timothy Keyes"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
description: > 
  Read this vignette to learn how to identify clusters of cells with shared
  phenotypic characteristics using {tidytof}.
vignette: >
  %\VignetteIndexEntry{Clustering}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE, warning = FALSE}
library(tidytof)
library(dplyr)
library(FlowSOM)
```

Often, clustering single-cell data to identify communities of cells with shared characteristics is a major goal of CyTOF data analysis.

To do this, `{tidytof}` provides the `tof_cluster()` verb. Several clustering methods are implemented in `{tidytof}`, including the following: 

- [FlowSOM](https://pubmed.ncbi.nlm.nih.gov/25573116/)
- [k-means](https://www.jstor.org/stable/2346830?origin=crossref&seq=1#metadata_info_tab_contents)
- [PhenoGraph](https://pubmed.ncbi.nlm.nih.gov/26095251/)
- [Supervised distance-based clustering](https://pubmed.ncbi.nlm.nih.gov/29505032/)
- [X-shift](https://pubmed.ncbi.nlm.nih.gov/27183440/)

Each of these methods are wrapped by `tof_cluster()`. 

## Clustering with `tof_cluster()`

To demonstrate, we can apply the FlowSOM clustering algorithm to `{tidytof}`'s built-in `phenograph_data`. Note that `phenograph_data` contains 3000 total cells (1000 each from 3 clusters identified in the [original PhenoGraph publication](https://pubmed.ncbi.nlm.nih.gov/26095251/)).

```{r}
phenograph_clusters <- 
  phenograph_data %>% 
  tof_preprocess() %>% 
  tof_cluster(cluster_cols = contains("cd"), method = "flowsom")

phenograph_clusters %>% 
  select(sample_name, .flowsom_metacluster, everything()) %>% 
  head()
```

The output of `tof_cluster` is a `tof_tbl` identical to the input tibble, but now with the addition of an additional column (in this case, ".flowsom_metacluster") that encodes the cluster id for each cell in the input `tof_tbl`. Note that all output columns added to a tibble or `tof_tbl` by `{tidytof}` begin with a full-stop (".") to reduce the likelihood of collisions with existing column names.

Because the output of `tof_cluster` is a `tof_tbl`, we can use `dplyr`'s `count` method to assess the accuracy of the FlowSOM clustering compared to the original clustering from the PhenoGraph paper.

```{r}
phenograph_clusters %>% 
  count(phenograph_cluster, .flowsom_metacluster, sort = TRUE)
```

Here, we can see that the FlowSOM algorithm groups most cells from the same PhenoGraph cluster with one another (with a small number of mistakes).

To change which clustering algorithm `tof_cluster` uses, alter the `method` flag.

```{r}
# use the phenograph algorithm 
phenograph_data %>% 
  tof_preprocess() %>% 
  tof_cluster(
    cluster_cols = contains("cd"),
    method = "phenograph"
  ) %>% 
  head()
```

To change the columns used to compute the clusters, change the `cluster_cols` flag. And finally, if you want to return a one-column `tibble` that only includes the cluster labels (as opposed to the cluster labels added as a new column to the input `tof_tbl`), set `augment` to `FALSE`.

```{r}
# will result in a tibble with only 1 column (the cluster labels)
phenograph_data %>% 
  tof_preprocess() %>% 
  tof_cluster(
    cluster_cols = contains("cd"),
    method = "kmeans", 
    augment = FALSE
  ) %>% 
  head()
```

## Metaclustering with `tof_metacluster()`

[Under Construction...]
