---
title: "Building predictive models"
author: "Timothy Keyes"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
description: > 
  Read this vignette to learn how to compute summary statistics like cluster 
  abundance and cluster marker expression using {tidytof} 
vignette: >
  %\VignetteIndexEntry{10. Modeling}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

```{r setup}
library(tidytof)
library(stringr)
```


`{tidytof}` implements several functions for building predictive models using sample- or patient-level data. 

## Accessing the data for this vignette

To illustrate how they work, first we download some patient-level data from [this paper](https://pubmed.ncbi.nlm.nih.gov/29505032/) and combine it with sample-level clinical annotations in one of `{tidytof}`'s built-in datasets (`ddpr_metadata`).


```{r}
data(ddpr_metadata)

# link for downloading the sample-level data from the Nature Medicine website
data_link <-
    "https://static-content.springer.com/esm/art%3A10.1038%2Fnm.4505/MediaObjects/41591_2018_BFnm4505_MOESM3_ESM.csv"

# download the data and combine it with clinical annotations
ddpr_patients <-
    readr::read_csv(data_link, skip = 2L, n_max = 78L, show_col_types = FALSE) |>
    dplyr::rename(patient_id = Patient_ID) |>
    dplyr::left_join(ddpr_metadata, by = "patient_id") |>
    dplyr::filter(!str_detect(patient_id, "Healthy"))

# preview only the metadata (i.e. non-numeric) columns
ddpr_patients |>
    dplyr::select(where(~ !is.numeric(.x))) |>
    head()
```

The data processing steps above result in a tibble called `ddpr_patients`. The numeric columns in `ddpr_patients` represent aggregated cell population features for each sample (see [Supplementary Table 5 in this paper](https://www.nature.com/articles/nm.4505#MOESM3) for details). The non-numeric columns represent clinical metadata about each sample (run `?ddpr_metadata` for more information). Of the metadata columns, the most important are the ones that indicate if a patient will develop refractory disease ("relapse"), and when/if that will happen. This information is stored in the `relapse_status` and `time_to_relapse` columns, respectively. 

There are also a few preprocessing steps that we might want to perform now to save us some headaches when we're fitting models later. 

```{r}
ddpr_patients <-
    ddpr_patients |>
    # convert the relapse_status variable to a factor
    # and create the time_to_event and event columns for survival modeling
    dplyr::mutate(
        relapse_status = as.factor(relapse_status),
        time_to_event = dplyr::if_else(relapse_status == "Yes", time_to_relapse, ccr),
        event = dplyr::if_else(relapse_status == "Yes", 1, 0)
    )
```

In the next part of this vignette, we'll use this patient-level data to build some 
predictive models using resampling procedures like k-fold cross-validation and 
bootstrapping. 

## Building a classifier using elastic net-regularized logistic regression

First, we can build an elastic net classifier to predict which patients will relapse and which patients won't (ignoring time-to-event data for now). For this, we can use the `relapse_status` column in `ddpr_patients` as the outcome variable: 

```{r}
# find how many of each outcome we have in our cohort
ddpr_patients |>
    dplyr::count(relapse_status)
```

We can see that not all of our samples are annotated, so we can throw away all the samples that don't have a clinical outcome associated with them. 

```{r}
ddpr_patients_unannotated <-
    ddpr_patients |>
    dplyr::filter(is.na(relapse_status))

ddpr_patients <-
    ddpr_patients |>
    dplyr::filter(!is.na(relapse_status))
```

In the original DDPR paper, 10-fold cross-validation was used to tune a glmnet 
model and to estimate the error of that model on new datasets. Here, we can use the `tof_split_data()` function to split our cohort into a training and test set either once 10 times using k-fold cross-validation or bootstrapping. Reading the documentation of `tof_split_data()` demonstrates how to use other resampling methods (like bootstrapping). 

```{r}
set.seed(3000L)

training_split <-
    ddpr_patients |>
    tof_split_data(
        split_method = "k-fold",
        num_cv_folds = 10,
        strata = relapse_status
    )

training_split
```

The output of `tof_split_data()` varies depending on which `split_method` is used. For cross-validation, the result is a `rset` object from the [rsample](https://rsample.tidymodels.org/) package. `rset` objects are a type of tibble with two columns: 

* `splits` - a column in which each entry is an `rsplit` object (which contains a single resample of the full dataset)
* `id` - a character column in which each entry represents the name of the fold that each entry in `splits` belongs to.

We can inspect one of the resamples in the `splits` column to see what they contain:

```{r}
my_resample <- training_split$splits[[1]]

print(my_resample)
```

Note that you can use `rsample::training` and `rsample::testing` to return the training and test observations from each resampling: 

```{r}
my_resample |>
    rsample::training() |>
    head()
```

```{r}
my_resample |>
    rsample::testing() |>
    head()
```

From here, we can feed  `training_split` into the `tof_train_model` function to [tune](https://www.tmwr.org/tuning.html) a logistic regression model that predicts the relapse_status of a leukemia patient. Be sure to check out the `tof_create_grid` documentation to learn how to make a hyperparameter search grid for model tuning (in this case, we limit the mixture parameter to a value of 1, which fits a sparse lasso model).

Also note that, in this case, for illustrative purposes we're only incorporating features from **one** of the populations of interest (population 2) into the model, whereas the original model incorporated features from all 12 populations (and likely required quite a bit of computational power as a result).

```{r, warning = FALSE}
hyperparams <- tof_create_grid(mixture_values = 1)

class_mod <-
    training_split |>
    tof_train_model(
        predictor_cols = c(contains("Pop2")),
        response_col = relapse_status,
        model_type = "two-class",
        hyperparameter_grid = hyperparams,
        impute_missing_predictors = TRUE,
        remove_zv_predictors = TRUE # often a smart decision
    )
```

The output of `tof_train_model` is a `tof_model`, an object containing information about the trained model (and that can be passed to the `tof_predict` and `tof_assess_model` verbs). When a `tof_model` is printed, some information about the optimal hyperparamters is printed, and so is a table of the nonzero model coefficients in the model. 

```{r}
print(class_mod)
```

After training our model, we might be interested in seeing how it performs. One way to assess a classification model is to see how well it works when applied directly back to the data it was trained on (the model's "training data"). To do so, we can use the `tof_assess_model()` function with no arguments:

```{r}
training_classifier_metrics <-
    class_mod |>
    tof_assess_model()
```

`tof_assess_model()` returns a list of several model assessment metrics that differ depending on the kind of `tof_model` you trained. For two-class classifier models, among the most useful of these is the `confusion_matrix`, which shows how your classifier classified each observation relative to its true class assignment.

```{r}
training_classifier_metrics$confusion_matrix
```

In this case, we can see that our model performed perfectly on its training data (which is expected, as the model was optimized using these data themselves!). You can also visualize the model's performance using the `tof_plot_model()` verb, which in the case of a two-class model will give us a Receiver-Operating Characteristic (ROC) curve: 

```{r}
class_mod |>
    tof_plot_model()
```

As shown above, `tof_plot_model()` will return a receiver-operating curve for a two-class model. While it's unusual to get an AUC of 1 in the machine learning world, we can note that in this case, our classification problem wasn't particularly difficult (and we had a lot of input features to work with).

After training a model, it generally isn't sufficient to evaluate how a model performs on the training data alone, as doing so will provide an overly-optimistic representation of how the model would perform on data it's never seen before (this problem is often called "overfitting" the model to the training data). To get a fairer estimate of our model's performance on new datasets, we can also evaluate its cross-validation error by calling `tof_assess_model()` and `tof_plot_model()` with the `new_data` argument set to "tuning".

```{r}
cv_classifier_metrics <-
    class_mod |>
    tof_assess_model(new_data = "tuning")

class_mod |>
    tof_plot_model(new_data = "tuning")
```

In this case, we plot an ROC Curve using the predictions for each observation when it was **excluded** from model training during cross-validation, an approach that gives a more accurate estimate of a model's performance on new data than simple evaluation on the training dataset.

## Building a survival model using elastic net-regularized cox regression

Building off of the ideas above, a more sophisticated way to model our data is not simply to predict who will relapse and who won't, but to build a time-to-event model that estimates patients' probabilities of relapse as a function of time since diagnosis. This approach is called "survival modeling" (specifically, in this case we use Cox-proportional hazards modeling) because it takes into account that patients have adverse events at different times during their course of disease (i.e. not everyone who relapses does so at the same time).

To build a survival model using `{tidytof}`, use the `tof_train_model()` verb while setting the `model_type` flag to "survival". In addition, you need to provide two outcome columns. The first of these columns (`event_col`) indicates if a patient relapsed (i.e. experienced the event-of-interest) or if they were censored after a certain amount of follow-up time. The second (`time_col`) indicates how much time it took for the patient to relapse or to be censored from the analysis.

```{r, warning = FALSE}
hyperparams <- tof_create_grid(mixture_values = 1)

survival_mod <-
    training_split |>
    tof_train_model(
        predictor_cols = c(contains("Pop2")),
        time_col = time_to_event,
        event_col = event,
        model_type = "survival",
        hyperparameter_grid = hyperparams,
        impute_missing_predictors = TRUE,
        remove_zv_predictors = TRUE # often a smart decision
    )

print(survival_mod)
```

Once a survival model is trained, it can be used to predict a patient's probability of the event-of-interest at different times post-diagnosis. However, the most common way that survival models are applied in practice is to use each patient's predicted relative risk of the event-of-interest to divide patients into low- and high-risk subgroups. `{tidytof}` can do this automatically according to the most optimal split obtained using the log-rank test of all possible split points in the dataset with `tof_assess_model()`. In addition, it will return the predicted survival curve for each patient over time: 

```{r}
survival_metrics <-
    survival_mod |>
    tof_assess_model()

survival_metrics
```

For survival models, `tof_plot_model()` plots the average survival curves for both the low- and high-risk groups: 

```{r}
survival_mod |>
    tof_plot_model()
```



```{r}
cv_survival_metrics <-
    survival_mod |>
    tof_assess_model(new_data = "tuning")
```


```{r}
survival_mod |>
    tof_plot_model(new_data = "tuning")
```



# Session info

```{r}
sessionInfo()
```





